---
title: "Sampling distribution of coefficient estimates in linear regression"
output: pdf_document
documentclass: extarticle
classoption: 14pt
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

 * We want to measure how useful a linear model is for predicting the response variable.
 * The residual standard error is the standard deviation of the residuals
    * Smaller residual standard error means predictions are better
 * The $R^2$ is the square of the correlation coefficient $r$
    * Larger $R^2$ means the model is better
    * Can also be interpreted as "proportion of variation in the response variable accounted for by the linear model" - see later statistics classes or the book for why.

## Example data set

We have a data set with observations of four variables measuring advertising budgets and sales for a product in each of 200 markets:

 * `sales` is a measure of sales volume in thousands of units
 * `TV` is TV advertising budget in thousands of dollars
 * `radio` is radio advertising budget in thousands of dollars
 * `newspaper` is newspaper advertising budget in thousands of dollars

Below are plots displaying three separate simple linear regression fits to the data.  In all three plots/models, `sales` is the response variable; the explanatory variable is different for each model.

```{r, fig.height = 3, warning=FALSE, message=FALSE, echo = FALSE}
library(readr)
library(ggplot2)
library(dplyr)
library(gridExtra) # for grid.arrange, puts plots next to each other

Advertising <- read_csv("http://www.evanlray.com/data/islr/Advertising.csv")

p1 <- ggplot(data = Advertising, mapping = aes(x = TV, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw(base_size = 14)

p2 <- ggplot(data = Advertising, mapping = aes(x = radio, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw(base_size = 14)

p3 <- ggplot(data = Advertising, mapping = aes(x = newspaper, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw(base_size = 14)

grid.arrange(p1, p2, p3, nrow = 1)
```

## Residual Standard Error

 * Residuals: $e_i = y_i - \hat{y}_i$ (vertical distance between point and line)
 * Smaller residuals mean the predictions were better.
 * Measure "size" of residuals with the standard deviation.  For reasons discussed later, call this the residual standard error.

```{r, fig.height = 3, warning=FALSE, message=FALSE, echo = FALSE}
library(readr)
library(ggplot2)
library(gridExtra) # for grid.arrange, puts plots next to each other

Advertising <- read_csv("http://www.evanlray.com/data/islr/Advertising.csv")

p1 <- ggplot(data = Advertising, mapping = aes(x = TV, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw(base_size = 14)

fit <- lm(sales ~ TV, data = Advertising)
Advertising <- Advertising %>%
  mutate(
    residual = residuals(fit)
  )
p4 <- ggplot(data = Advertising, mapping = aes(x = residual)) + geom_density() + xlim(c(-15, 15)) +
  theme_bw(base_size = 14)

p2 <- ggplot(data = Advertising, mapping = aes(x = radio, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw(base_size = 14)

fit <- lm(sales ~ radio, data = Advertising)
Advertising <- Advertising %>%
  mutate(
    residual = residuals(fit)
  )
p5 <- ggplot(data = Advertising, mapping = aes(x = residual)) + geom_density() + xlim(c(-15, 15)) +
  theme_bw(base_size = 14)

p3 <- ggplot(data = Advertising, mapping = aes(x = newspaper, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw(base_size = 14)

fit <- lm(sales ~ newspaper, data = Advertising)
Advertising <- Advertising %>%
  mutate(
    residual = residuals(fit)
  )
p6 <- ggplot(data = Advertising, mapping = aes(x = residual)) + geom_density() + xlim(c(-15, 15)) +
  theme_bw(base_size = 14)

grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 2)
```

Residual standard errors:

 * For regression on TV: 3.26
 * For regression on radio: 4.28
 * For regression on newspaper: 5.09

Recall that:

 * If a variable follows an approximately normal distribution, about 95% of observations are within 2 standard deviations of the mean.
 * The mean of the residuals is 0.

### 1. What is the interpretation of the residual standard deviation for the regression using TV as the explanatory variable, based on the "2 standard deviations rule"?

\newpage

### 2. What is the interpretation of the residual standard deviation for the regression using radio as the explanatory variable, based on the "2 standard deviations rule"?

\vspace{4cm}

### 3. What is the interpretation of the residual standard deviation for the regression using newspaper as the explanatory variable, based on the "2 standard deviations rule"?

\vspace{4cm}

## $R^2$

Square of the correlation coefficient $r$: between 0 and 1, closer to 1 is better.

 * For regression on TV: 0.61
 * For regression on radio: 0.33
 * For regression on newspaper: 0.05

```{r}
Advertising %>% select(TV, sales) %>% cor()
(0.7822244)^2
```

## Finding things in the R output

```{r}
fit <- lm(sales ~ TV, data = Advertising)
summary(fit)
```
