---
title: "R Commands for Inference for Proportions"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
documentclass: extarticle
classoption: 14pt
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
require(ggplot2)
require(scales)
require(dplyr)
require(tidyr)
require(readr)
require(mosaic)
```

\vspace{-0.5cm}

Suppose we take a sample of size $n = 40$ and our null hypothesis is $H_0: p = 0.2$

If the null hypothesis is true and conditions check out, our sampling distribution is $X \sim \text{Binomial}(40, 0.2)$ 

There are 3 options for what $H_A$ could be:

\vspace{-0.7cm}

#### Option 1: $H_A: p > 0.2$ (do not use this confidence interval)

Suppose that we observe $x = 14$.  The p-value is $P(X \geq 14) = P(X > 13)$.

```{r, echo = TRUE}
pbinom(q = 14 - 1, size = 40, prob = 0.2, lower.tail = FALSE)
```

```{r, echo = TRUE, eval = FALSE}
binom.test(x = 14, n = 40, p = 0.2, alternative = "greater")
```
\includegraphics{ss_greater}

```{r, echo = FALSE, fig.height=1.8, fig.width=8}
x <- 14
n <- 40
p <- 0.2
Paul_success_probs <- data.frame(
  num_successes = seq(from = 0, to = n),
  pv = factor(c(rep(0, x), rep(1, n - x + 1))),
  probability = dbinom(x = seq(from = 0, to = n), size = n, prob = p))

ggplot() +
  geom_col(mapping = aes(x = num_successes, y = probability, fill = pv),
    data = Paul_success_probs) +
  geom_vline(xintercept = x) +
  xlab("Number of Successes") +
  scale_fill_manual("Included\nin p-value\ncalculation?", labels = c("No", "Yes"), values = c("black", "red")) +
  theme_gray(base_size = 14)
```

#### Option 2: $H_A: p < 0.2$ (do not use this confidence interval)

Suppose that we observe $x = 6$.  The p-value is $P(X \leq 6)$.

```{r, echo = TRUE}
pbinom(q = 6, size = 40, prob = 0.2)
```

```{r, echo = TRUE, eval = FALSE}
binom.test(x = 6, n = 40, p = 0.2, alternative = "less")
```
\includegraphics{ss_less}

```{r, echo = FALSE, fig.height=1.8, fig.width=8}
x <- 6
n <- 40
p <- 0.2
Paul_success_probs <- data.frame(
  num_successes = seq(from = 0, to = n),
  pv = factor(c(rep(1, x + 1), rep(0, n - x))),
  probability = dbinom(x = seq(from = 0, to = n), size = n, prob = p))

ggplot() +
  geom_col(mapping = aes(x = num_successes, y = probability, fill = pv),
    data = Paul_success_probs) +
  geom_vline(xintercept = x) +
  xlab("Number of Successes") +
  scale_fill_manual("Included\nin p-value\ncalculation?", labels = c("No", "Yes"), values = c("black", "red")) +
  theme_gray(base_size = 14)
```

#### Option 3: $p \neq 0.2$, Confidence Interval

Suppose we observe $x = 14$.  The p-value is the probability of getting a test statistic at least as far from the expected result if the null hypothesis was true.  We will only use the `binom.test` function for p-value calculations in this case.

```{r, echo = TRUE, eval = FALSE}
binom.test(x = 14, n = 40, p = 0.2, alternative = "two.sided",
  conf.level = 0.95)
```

\includegraphics{ss_notequal}

```{r, echo = FALSE, fig.height=1.8, fig.width=8}
x <- 14
n <- 40
p <- 0.2

expected <- n * p
diff <- abs(x - expected)

Paul_success_probs <- data.frame(
  num_successes = seq(from = 0, to = n),
  pv = factor(c(rep(1, expected - diff + 1), rep(0, 2 * diff - 1), rep(1, n - 2 * diff - 1))),
  probability = dbinom(x = seq(from = 0, to = n), size = n, prob = p))

ggplot() +
  geom_col(mapping = aes(x = num_successes, y = probability, fill = pv),
    data = Paul_success_probs) +
  geom_vline(xintercept = x) +
  xlab("Number of Successes") +
  scale_fill_manual("Included\nin p-value\ncalculation?", labels = c("No", "Yes"), values = c("black", "red")) +
  theme_gray(base_size = 14)
```


