---
title: "Stat 140 - Quiz 1 Sample"
output: pdf_document
header-includes:
   - \usepackage{soul}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(dplyr)
library(ggplot2)
```

## What's Your Name? __________________________________

## Which section are you in? _______________________________

\textbf{This is a sample quiz.  For the real quiz, I will use a different data set, but will ask the questions below about the new data set with minimal modification.}

Below are the first few rows of a data frame named \texttt{NHANES}.  NHANES stands for "National Health and Nutrition Examination Surveys", and the data frame contains information about the health of randomly sampled Americans.

```{r, load_data, message = FALSE, echo = FALSE}
library(NHANES)
library(dplyr)
library(tidyr)
library(ggplot2)
library(mosaic)

NHANES <- NHANES[, c(1, 3, 4, 17, 20, 21, 25, 26, 40)]
NHANES <- as.data.frame(NHANES)
NHANES <- distinct(NHANES)
```

```{r, message=FALSE, echo = FALSE}
head(NHANES)
```

### 1. What is each observational unit in this data set?

\vspace{2cm}

### 2. For each of the following variables, is that variable categorical or quantitative?  If it is categorical, is it ordinal or nominal?

* `Gender`
* `Height`
* `Diabetes`

### 3. The following command counts how many observational units are in each combination of levels of the gender and diabetes variables.

```{r, echo = FALSE}
NHANES_orig <- NHANES
NHANES <- NHANES %>% filter(!is.na(Diabetes))
```

```{r}
NHANES %>%
  count(Diabetes, Gender) %>%
  spread(Gender, n)
```

```{r, echo = FALSE}
NHANES <- NHANES_orig
```

#### a. Calculate the joint distribution of Diabetes and Gender

\vspace{3cm}

#### b. Calculate the marginal distribution of Diabetes

\vspace{3cm}

#### c. Calculate the conditional distribution of Diabetes given that the subject's Gender is male

\vspace{3cm}

#### d. Calculate the conditional distribution of Diabetes given that the subject's Gender is female

\vspace{3cm}

#### e. Is a person's Diabetes status independent of their Gender?

