---
title: "Two Lines: Crocodiles!!"
output:
  pdf_document:
    keep_tex: true
header-includes:
   - \usepackage{soul}
   - \usepackage{booktabs}
documentclass: extarticle
classoption: 14pt
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
require(readr)
```

```{r, message = FALSE, echo = FALSE}
# set seed and sample_frac to get both species in the first few rows.
set.seed(123)

crocs <- read_csv("http://www.evanlray.com/data/sdm4/Crocodile_lengths.csv") %>%
  mutate(Species = factor(Species)) %>%
  sample_frac(size = 1) %>%
  as.data.frame()
```

```{r}
head(crocs)
```

```{r, fig.height=2}
ggplot(data = crocs, mapping = aes(x = Body, y = Head, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

### 2 lines by filtering to create separate data sets

```{r}
aus_crocs <- crocs %>% filter(Species == "Australian")
aus_fit <- lm(Head ~ Body, data = aus_crocs)
summary(aus_fit)
```

```{r}
ind_crocs <- crocs %>% filter(Species == "Indian")
ind_fit <- lm(Head ~ Body, data = ind_crocs)
summary(ind_fit)
```

#### Questions we'd like to be able to answer (but can't with this output):

1. Is there statistically significant evidence that the intercepts for these lines are different?
1. Is there statistically significant evidence that the slopes for these lines are different?

\newpage

### 2 parallel lines (same slope)

* Our Goal: Equations for two lines
\begin{align*}
\text{Predicted Head Length for Australian Crocs } &= \hat{\beta}_0^{Australian} + \hat{\beta}_{1} \times (\text{Body Length}) \\
\text{Predicted Head Length for Indian Crocs } &= \hat{\beta}_0^{Indian} + \hat{\beta}_{1} \times (\text{Body Length})
\end{align*}

 * Note: Different intercepts, same slope.

```{r, echo = TRUE}
parallel_lines_fit <- lm(Head ~ Body + Species, data = crocs)
summary(parallel_lines_fit)
```

* R gives us a single combined equation:
$$\text{Predicted Head Length} = \hat{\beta}_0 + \hat{\beta}_{1} (\text{Body Length}) + \hat{\beta}_2 \text{SpeciesIndian}$$

$$\text{Predicted Head Length} = 2.27 + 0.13 (\text{Body Length}) + 8.89 \text{SpeciesIndian}$$

\newpage

#### What is the `SpeciesIndian` variable?

* Behind the scenes, R creates a new **indicator variable** called `SpeciesIndian`:
$$\text{{\tt SpeciesIndian}} = \begin{cases} 1 & \text{ if the species for crocodile $i$ is Indian.} \\
0 & \text{ otherwise (in this case, the species is Australian)} \end{cases}
$$

* R doesn't modify the data frame (it creates a secret copy in the background), but it would look like this:

```{r, echo = FALSE}
crocs <- mutate(crocs,
  SpeciesIndian = as.integer(Species == "Indian"))
```

```{r, echo = TRUE}
head(crocs)
```

Above, we obtained this estimated equation:

$$\text{Predicted Head Length} = 2.27 + 0.13 (\text{Body Length}) + 8.89 \text{SpeciesIndian}$$

#### What is the estimated equation describing the relationship between body length and head length, for Australian crocodiles?

\vspace{3cm}

#### What is the estimated equation describing the relationship between body length and head length, for Indian crocodiles?

\vspace{3cm}

#### What is the interpretation of $\widehat{\beta}_0 = 2.27$?

\vspace{3cm}

#### What is the interpretation of $\widehat{\beta}_1 = 0.13$?

\vspace{3cm}

#### What is the interpretation of $\widehat{\beta}_2 = 8.89$?

\vspace{3cm}

#### Using the output from the summary function, conduct a test of the claim that the intercept of the line describing the relationship between body length and head length in the population of all Australian crocodiles is the same as the intercept of the line describing the relationship between body length and head length in the population of all Indian crocodiles.

\newpage

### Lines with different slopes (interactions)

To get different slopes, use `Body * Species` instead of `Body + Species`:

```{r, echo = TRUE}
two_lines_fit <- lm(Head ~ Species * Body, data = crocs)
summary(two_lines_fit)
```

#### What is the estimated equation from this model?

\vspace{3cm}

#### What is the estimated equation describing the relationship between body length and head length, for Australian crocodiles?

\vspace{3cm}

#### What is the estimated equation describing the relationship between body length and head length, for Indian crocodiles?

\vspace{3cm}

#### What is the interpretation of $\widehat{\beta}_0 = 3.463$?

\vspace{3cm}

#### What is the interpretation of $\widehat{\beta}_1 = 7.075$?

\vspace{3cm}

#### What is the interpretation of $\widehat{\beta}_2 = 0.125$?

\vspace{3cm}

#### What is the interpretation of $\widehat{\beta}_3 = 0.006$?

\newpage

#### Using the output from the summary function, conduct a test of the claim that the *slope* of the line describing the relationship between body length and head length in the population of all Australian crocodiles is the same as the *slope* of the line describing the relationship between body length and head length in the population of all Indian crocodiles.

