---
title: "Regression -- Two Lines"
author: "Evan L. Ray"
date: "October 11, 2017"
output: ioslides_presentation
---

<script type="text/x-mathjax-config"> 
    MathJax.Hub.Config({ 
        "HTML-CSS": { scale: 90, linebreaks: { automatic: true } }, 
        SVG: { linebreaks: { automatic:true } }, 
        displayAlign: "left" });
</script>

<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
  .col3 {
    columns: 3 100px;
    -webkit-columns: 3 100px;
    -moz-columns: 3 100px;
  }
  .MathJax {
    font-size: 0.9em;
  }
</style>

<script type="text/x-mathjax-config">
  MathJax.Hub.Config({ TeX: { extensions: ["color.js"] }});
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(ggplot2)
require(dplyr)
require(tidyr)
require(readr)
```

## Crocodiles!!

* Measurements of
    * length of crocodile's body in cm
    * length of crocodile's head in cm
    * crocodile species (Australian or Indian)

```{r, echo = FALSE, message = FALSE}
require(readr)

set.seed(123)
crocs <- read_csv("https://mhc-stat140-2017.github.io/data/sdm4/Crocodile_lengths.csv") %>%
  mutate(Species = factor(Species, levels = c("Australian", "Indian"))) %>%
  sample_frac(size = 1) %>%
  as.data.frame()
```

```{r, echo = TRUE}
head(crocs)
```

## A first plot

```{r, echo = TRUE, fig.height=3, fig.width=8}
ggplot() +
  geom_point(mapping = aes(x = Body, y = Head),
    data = crocs) +
  geom_smooth(mapping = aes(x = Body, y = Head),
    data = crocs,
    method = "lm",
    se = FALSE)
```

## Residuals Density Plot

```{r, echo = FALSE}
one_line_fit <- lm(Head ~ Body, data = crocs)
crocs2 <- mutate(crocs,
  residual = residuals(one_line_fit))
ggplot() +
  geom_density(mapping = aes(x = residual), data = crocs2)
```

## A second plot

```{r, echo = TRUE, fig.height=3, fig.width=8}
ggplot() +
  geom_point(mapping = aes(x = Body, y = Head, color = Species),
    data = crocs) +
  geom_smooth(mapping = aes(x = Body, y = Head, color = Species),
    data = crocs,
    method = "lm",
    se = FALSE)
```

## Residuals Density Plot -- Two Lines

```{r, echo = FALSE}
two_lines_fit <- lm(Head ~ Species * Body, data = crocs)
crocs2 <- mutate(crocs,
  residual = residuals(two_lines_fit))
ggplot() +
  geom_density(mapping = aes(x = residual), data = crocs2)
```


## Fitting Two Parallel Lines (Same Slope)

* Our Goal: Equations for two lines
    * Predicted Head Length for Australian Crocs  
    $\qquad = b_0^{Australian}$ + $b_{1}$ (Body Length)
    * Predicted Head Length for Indian Crocs  
    $\qquad = b_0^{Indian}$ + $b_{1}$ (Body Length)

```{r, echo = TRUE}
parallel_lines_fit <- lm(Head ~ Species + Body, data = crocs)
coef(parallel_lines_fit)
```

* R gives us a single combined equation:
    * Predicted Head Length =  
    $\qquad$ 2.27 + 8.89 SpeciesIndian + 0.13 (Body Length)


## Indicator Variables

* Behind the scenes, R creates a new **indicator variable** called `SpeciesIndian`.
    * Equals 1 if the species is "Indian"
    * Equals 0 otherwise (in this case, that means the species is "Australian")
* R doesn't modify the data frame, but it would look like this:
```{r, echo = FALSE}
crocs <- mutate(crocs,
  SpeciesIndian = as.integer(Species == "Indian"))
```

```{r, echo = TRUE}
head(crocs)
```


## Back to the Equations

* Two slides ago, we got this equation from R:
    * Predicted Head Length  
    $\qquad$ = 2.27 + 8.89 `SpeciesIndian` + 0.13 (Body Length)

* For an Australian crocodile, `SpeciesIndian` = 0, so we get
    * Predicted Head Length  
    $\qquad$ = 2.27 + 0.13 (Body Length)

* For an Indian crocodile, `SpeciesIndian` = 1, so we get
    * Predicted Head Length  
    $\qquad$ = 2.27 + 8.89 + 0.13 (Body Length)  
    $\qquad$ = 11.16 + 0.13 (Body Length)

## Interpretation of Coefficients

* Predicted Head Length  
  $\qquad$ = 2.27 + 8.89 `SpeciesIndian` + 0.13 (Body Length)

* Interpretation of intercept: The predicted head length for an Australian crocodile with a body length of 0cm is 2.27cm.

* Interpretation of the coefficient for `SpeciesIndian`: For a given body length, the predicted head length for an Indian crocodile is 8.89 cm longer than the predicted head length for an Australian crocodile.

* Interpretation of coefficient for body length is as usual: for each 1cm increase in body length, predicted head length increases by 0.13 cm.

## Two Lines with Different Slopes

* In the formula for `lm`, we use `Species * Body` instead of `Species + Body`:

```{r, echo = TRUE}
two_lines_fit <- lm(Head ~ Species * Body, data = crocs)
coef(two_lines_fit)
```

* Combined equation:
    * Predicted Head Length =  
    $\qquad$ 3.463 + 7.075 `SpeciesIndian` +  
    $\qquad$ 0.125 `Body` + 0.006 `SpeciesIndian` * `Body`


## Separate Equations for Each Species

* Combined equation:
    * Predicted Head Length =  
    $\qquad$ 3.463 + 7.075 `SpeciesIndian` +  
    $\qquad$ 0.125 `Body` + 0.006 `SpeciesIndian` * `Body`

* For an Australian crocodile, `SpeciesIndian` = 0, so we get
    * Predicted Head Length  
    $\qquad$ = 3.463 + 0.125 * `Body`

* For an Indian crocodile, `SpeciesIndian` = 1, so we get
    * Predicted Head Length  
    $\qquad$ = 3.463 + 7.075 + 0.125 * `Body` + 0.006 * `Body`  
    $\qquad$ = 10.538 + 0.131 * `Body`

## Interpretations of Coefficients from R

* Predicted Head Length = 3.463 + 7.075 `SpeciesIndian` +  
    $\qquad$ 0.125 `Body` + 0.006 `SpeciesIndian` * `Body`

* Intercept: The predicted head length for an Australian crocodile with a body length of 0cm is 3.463cm.

* `SpeciesIndian`: The predicted head length for an Indian crocodile with a body length of 0cm is 7.075cm larger than the predicted head length for an Australian crocodile with a body length of 0cm.

* `Body`: For each 1cm increase in body length of an Australian crocodile, the predicted head length increases by 0.13 cm.

* `SpeciesIndian` * `Body`: For each 1cm increase in body length, the predicted head length of an Indian crocodile increases by 0.006cm more than the predicted head length of an Australian crocodile.
