---
title: "The Normal Distribution (Chapter 5)"
output: pdf_document
documentclass: extarticle
classoption: 14pt
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(ggplot2)
library(grid)
require(dplyr)
require(tidyr)
require(readr)
require(mosaic)
```

Two examples (black: observed sample data; red: a normal model with the same mean and standard deviation)

```{r, echo = FALSE, message = FALSE, fig.height=3.5}
car_speeds <- read_csv("http://www.evanlray.com/data/sdm3/Chapter_06/Ch06_Car_speeds.csv")
colnames(car_speeds) <- "speed"
normal_mean <- mean(car_speeds$speed)
normal_sd <- sd(car_speeds$speed)

p1 <- ggplot() +
  geom_density(mapping = aes(x = speed), data=car_speeds) +
  stat_function(mapping = aes(x = speed),
    fun = dnorm,
    colour = "red",
    args = list(mean = normal_mean, sd = normal_sd),
    data = car_speeds) +
#  scale_x_continuous(
#    breaks = c(16, 31, normal_mean + seq(from = -1, to = 1)*normal_sd),
#    labels = c(16, 31, expression(paste(mu, " - ", sigma)), expression(paste(mu)), expression(paste(mu, " + ", sigma)))) +
  ggtitle("Car speeds in a 20 MPH zone") +
  theme_gray(base_size = 14)

pizza <- read_csv("http://www.evanlray.com/data/sdm3/Chapter_04/Ch04_Pizza_Prices.csv")
normal_mean <- mean(pizza$Price)
normal_sd <- sd(pizza$Price)

p3 <- ggplot() +
  geom_density(mapping = aes(x = Price), data=pizza) +
  stat_function(mapping = aes(x = Price),
    fun = dnorm,
    colour = "red",
    args = list(mean = normal_mean, sd = normal_sd),
    data = pizza) +
#  scale_x_continuous(
#    breaks = c(2.3, 2.9, normal_mean + seq(from = -1, to = 1)*normal_sd),
#    labels = c(2.3, 2.9, expression(paste(mu, " - ", sigma)), expression(paste(mu)), expression(paste(mu, " + ", sigma)))) +
  ggtitle("Prices of plain pizza slices in Dallas, TX") +
  theme_gray(base_size = 14)

grid.newpage()
#pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 2)))
pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 1)))
#pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 2)))
print(p1, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
#print(p2, vp = viewport(layout.pos.row = 1, layout.pos.col = 2))
print(p3, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
#print(p4, vp = viewport(layout.pos.row = 2, layout.pos.col = 2))
```

Let the random variable $X$ be the numeric value of one of these variables for a randomly sampled item from the population

 * Example: $X$ is the speed of a randomly selected car in a 20 MPH speed zone.
 * Example: $X$ is the price of a piece of cheese pizza from a randomly selected restaurant in Dallas, TX.

We could **model** the value of $X$ as being a draw from a normal distribution

 * $X \sim \text{Normal}(\mu, \sigma)$
 * Read: X follows a normal distribution with mean $\mu$ ("mew", like a cat) and standard deviation $\sigma$ ("sigma")
 * $\mu$ and $\sigma$ determine the center and spread of the distribution

```{r, echo = FALSE, fig.height=2}
x_grid <- seq(from = -6, to = 6, by = 0.01)
n_grid <- length(x_grid)

mu1 <- 0
sigma1 <- 1

mu2 <- 1
sigma2 <- 0.2

mu3 <- -2
sigma3 <- 2

plot_df <- data.frame(
  x = rep(x_grid, 3),
  density = c(dnorm(x_grid, mean = mu1, sd = sigma1), dnorm(x_grid, mean = mu2, sd = sigma2), dnorm(x_grid, mean = mu3, sd = sigma3)),
  parameters = c(rep("mu = 0, sigma = 1", n_grid), rep("mu = 1, sigma = 0.2", n_grid), rep("mu = -2, sigma = 2", n_grid))
)

ggplot() +
  geom_line(aes(x = x, y = density, color = parameters), data = plot_df) +
  scale_color_manual("Mean, Standard Deviation",
    labels = c(
      expression(paste(mu, " = -2, ", sigma, " = 2")),
      expression(paste(mu, " = 0, ", sigma, " = 1")),
      expression(paste(mu, " = 1, ", sigma, " = 0.2"))),
      values = c("cornflowerblue", "orange", "purple")) +
  theme_gray(base_size = 14)
```

\newpage

Here are pictures of normal models for car speeds and pizza prices:

```{r, echo = FALSE, fig.height = 4}
normal_mean <- 23.84
normal_sd <- 3.56
plot_df <- data.frame(
  x = seq(from = normal_mean - 4 * normal_sd,
    to = normal_mean + 4 * normal_sd,
    length = 101)
)
plot_df1 <- data.frame(
  x = c(normal_mean - 1 * normal_sd,
    seq(from = normal_mean - 1 * normal_sd,
      to = normal_mean + 1 * normal_sd,
      length = 101),
    normal_mean + 1 * normal_sd
    ),
  density = c(0,
    dnorm(seq(from = normal_mean - 1 * normal_sd,
        to = normal_mean + 1 * normal_sd,
        length = 101),
      mean = normal_mean,
      sd = normal_sd),
    0)
  )

plot_df2 <- data.frame(
  x = c(normal_mean - 2 * normal_sd,
    seq(from = normal_mean - 2 * normal_sd,
      to = normal_mean + 2 * normal_sd,
      length = 101),
    normal_mean + 2 * normal_sd
    ),
  density = c(0,
    dnorm(seq(from = normal_mean - 2 * normal_sd,
        to = normal_mean + 2 * normal_sd,
        length = 101),
      mean = normal_mean,
      sd = normal_sd),
    0)
  )

p1 <- ggplot() +
  geom_polygon(aes(x = x, y = density), fill = "blue", alpha = 0.4, data = plot_df2) +
  geom_polygon(aes(x = x, y = density), fill = "orange", alpha = 0.8, data = plot_df1) +
  stat_function(mapping = aes(x = x),
    fun = dnorm,
    args = list(mean = normal_mean, sd = normal_sd),
    data = plot_df) +
  geom_vline(xintercept = normal_mean - 3*normal_sd) +
  geom_vline(xintercept = normal_mean - 2*normal_sd) +
  geom_vline(xintercept = normal_mean - 1*normal_sd) +
  geom_vline(xintercept = normal_mean) +
  geom_vline(xintercept = normal_mean + 1*normal_sd) +
  geom_vline(xintercept = normal_mean + 2*normal_sd) +
  geom_vline(xintercept = normal_mean + 3*normal_sd) +
  scale_x_continuous(
    breaks = c(normal_mean + seq(from = -3, to = 3)*normal_sd),
    labels = c(round(normal_mean + seq(from = -3, to = 3)*normal_sd, 2))) +
  ggtitle("Normal(23.84, 3.56) Model for Car Speeds") +
  theme_gray(base_size = 14)

normal_mean <- 2.62
normal_sd <- 0.16
plot_df <- data.frame(
  x = seq(from = normal_mean - 4 * normal_sd,
    to = normal_mean + 4 * normal_sd,
    length = 101)
)
plot_df1 <- data.frame(
  x = c(normal_mean - 1 * normal_sd,
    seq(from = normal_mean - 1 * normal_sd,
      to = normal_mean + 1 * normal_sd,
      length = 101),
    normal_mean + 1 * normal_sd
    ),
  density = c(0,
    dnorm(seq(from = normal_mean - 1 * normal_sd,
        to = normal_mean + 1 * normal_sd,
        length = 101),
      mean = normal_mean,
      sd = normal_sd),
    0)
  )

plot_df2 <- data.frame(
  x = c(normal_mean - 2 * normal_sd,
    seq(from = normal_mean - 2 * normal_sd,
      to = normal_mean + 2 * normal_sd,
      length = 101),
    normal_mean + 2 * normal_sd
    ),
  density = c(0,
    dnorm(seq(from = normal_mean - 2 * normal_sd,
        to = normal_mean + 2 * normal_sd,
        length = 101),
      mean = normal_mean,
      sd = normal_sd),
    0)
  )

p3 <- ggplot() +
  geom_polygon(aes(x = x, y = density), fill = "blue", alpha = 0.4, data = plot_df2) +
  geom_polygon(aes(x = x, y = density), fill = "orange", alpha = 0.8, data = plot_df1) +
  stat_function(mapping = aes(x = x),
    fun = dnorm,
    args = list(mean = normal_mean, sd = normal_sd),
    data = plot_df) +
  geom_vline(xintercept = normal_mean - 3*normal_sd) +
  geom_vline(xintercept = normal_mean - 2*normal_sd) +
  geom_vline(xintercept = normal_mean - 1*normal_sd) +
  geom_vline(xintercept = normal_mean) +
  geom_vline(xintercept = normal_mean + 1*normal_sd) +
  geom_vline(xintercept = normal_mean + 2*normal_sd) +
  geom_vline(xintercept = normal_mean + 3*normal_sd) +
  scale_x_continuous(
    breaks = c(normal_mean + seq(from = -3, to = 3)*normal_sd),
    labels = c(round(normal_mean + seq(from = -3, to = 3)*normal_sd, 2))) +
  ggtitle("Normal(2.62, 0.16) Model for Pizza Prices") +
  theme_gray(base_size = 14)

grid.newpage()
pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 1)))
print(p1, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p3, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
#print(p1 + xlim(c(0, 38)) + ylim(c(0, 2.5)), vp = viewport(layout.pos.row = 1, layout.pos.col = 2))
#print(p3 + xlim(c(0, 38)) + ylim(c(0, 2.5)), vp = viewport(layout.pos.row = 2, layout.pos.col = 2))
```

* Area under the curve is the probability of getting an observation in that region
* For any normal distribution,
    * 68% of observations are within $\mu \pm \sigma$ (orange area is about 0.68)
    * 95% of observations are within $\mu \pm 2 \sigma$ (sum of orange and blue areas is about 0.95)
    * Total area under the curve is 1 (all observations have some value of x).

### $z$-scores

For calculating probabilities, what matters is **how many standard deviations away from the mean a particular number $x$ is**.

This is the $z$-score of $x$: $z = \frac{x - \mu}{\sigma}$

Note:  If $X \sim \text{Normal}(\mu, \sigma)$, then $Z \sim \text{Normal}(0, 1)$

#### Example: Suppose $X$ is the speed of a randomly selected car in a 20MPH zone, and the distribution of speeds of such cars is $\text{Normal}(23.84, 3.56)$.  How many standard deviations above or below the mean is a car driving 27.4 miles per hour?

\vspace{2cm}

### Probability calculations in R

The `pnorm` function calculates probabilities involving the normal distribution.

We will provide the $z$-score as an argument to `pnorm`.

#### Example: What is the probability that a randomly selected car will be driving less than 27.4 miles per hour, assuming that car speeds follow a $\text{Normal}(23.84, 3.56)$ distribution?

The number I provided to `pnorm` below is the z-score of 27.4.

```{r, echo = TRUE}
pnorm(1)
```

\vspace{2cm}

#### Example: What is the probability that a randomly selected car will be driving more than 27.4 miles per hour, assuming that car speeds follow a $\text{Normal}(23.84, 3.56)$ distribution?

The number I provided to `pnorm` below is the z-score of 27.4.

```{r, echo = TRUE}
pnorm(1, lower.tail = FALSE)
```

\vspace{2cm}

#### Example: What is the probability that a randomly selected car will be driving exactly 27.4 miles per hour, assuming that car speeds follow a $\text{Normal}(23.84, 3.56)$ distribution?

\newpage

There is space to do calculations for the following examples in Lab 12 on Gryd.

#### Example: Suppose that the price of a slice of cheese pizza from a randomly selected restaurant in Dallas, TX follows a $\text{Normal}(2.62, 0.16)$ distribution.  Find the probability that a slice of pizza from a randomly selected restaurant costs less than $2.25.

\vspace{5cm}

#### Example: Suppose that the price of a slice of cheese pizza from a randomly selected restaurant in Dallas, TX follows a $\text{Normal}(2.62, 0.16)$ distribution.  Find the probability that a slice of pizza from a randomly selected restaurant costs more than $3.00.

\vspace{2cm}

