---
title: "Summarizing the Center and Spread of Quantitative Variables"
author: ''
date: "September 17, 2018"
output:
  pdf_document:
    keep_tex: true
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
   - \usepackage{booktabs}
documentclass: extarticle
classoption: 14pt
geometry: margin=0.6in
---


```{r, include=FALSE, message = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(readr)
library(dplyr)
require(mosaic)
library(grid)
```

```{r, include=FALSE, message = FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)

require(fivethirtyeight)                
```

# Summaries of Center (what is a "typical" value?)

Reminder of definitions from your reading:

Suppose we observe $n$ numbers, $x_1, \ldots, x_n$.

There are two commonly used statistics used to summarize the \textbf{center} of the distribution of these values:

 * The \textbf{mean} is the average of these values (add them up and divide by $n$).  We use $\bar{x}$ to denote the mean:

$$\bar{x} = \frac{\sum_{i = 1}^n x_i}{n} = \frac{x_1 + \cdots + x_n}{n}$$

 * The \textbf{median} is the middle value when you arrange them in order.  (If the sample size $n$ is even, you take the average of the middle two values)


## The situation:

It's 2013, and 6 friends are hanging out at their local bar.  Their incomes are \$30,000, \$32,000, \$34,000, \$36,000, \$38,000, and \$40,000.

### What is their mean income?

\vspace{2cm}

### What is their median income?

\newpage

## In walks BILL GATES!

According to the internet, in 2013 Bill Gates had an income of \$11.5 billion (in case you're curious, that works out to \$23,148 per minute).

### What is the mean income of the 6 friends and Bill Gates?  (Note that if you write it out with all the zeros, 11.5 billion is 11500000000; there are 8 zeros)

\vspace{3cm}

### What is the median income of the 6 friends and Bill Gates?

\vspace{3cm}

\newpage

# Summaries of Spread (how spread out are the values?)

There are three common measures of the **spread** of a distribution (how "wide" is it?):

1. We saw the **inter-quartile range** (IQR) last class:

IQR = Q3 - Q1 = 75th percentile - 25th percentile

The IQR is the width of an interval covering the middle half of the data.

2. The **variance** is (almost) the average squared difference of each observation from the mean.

$$\text{Variance} = \frac{\sum_{i = 1}^n (x_i - \bar{x})^2}{n - 1} = \frac{(x_1 - \bar{x})^2 + (x_2 - \bar{x})^2 + \cdots + (x_n - \bar{x})^2}{n - 1}$$

3. The **standard deviation** is the square root of the variance.  Intuitively, you can think of it as the average distance of the data points from the mean (although technically, that's not exactly right).

$$\text{Standard Deviation} = \sqrt{\text{Variance}} = \sqrt{\frac{\sum_{i = 1}^n (x_i - \bar{x})^2}{n - 1}}$$

### Let's use R to calculate these, rather than doing it by hand.

I have set up two different data frames - one with the numbers for just the friends, and a second including both the friends and Bill Gates.

```{r, echo = FALSE}
friends <- data.frame(
  person_number = 1:6,
  salary = c(30000, 32000, 34000, 36000, 38000, 40000)
)

friends_and_bill <- data.frame(
  person_number = 1:7,
  salary = c(30000, 32000, 34000, 36000, 38000, 40000, 11500000000)
)
```

#### Here are summaries of center and spread, including just the friends:

```{r}
friends
```

```{r}
friends %>%
  summarize(
    mean_salary = mean(salary),
    median_salary = median(salary),
    iqr_salary = IQR(salary),
    var_salary = var(salary),
    sd_salary = sd(salary)
  )
```

#### Here are summaries of center and spread, including Bill too:

```{r}
friends_and_bill
```

```{r}
friends_and_bill %>%
  summarize(
    mean_salary = mean(salary),
    median_salary = median(salary),
    iqr_salary = IQR(salary),
    var_salary = var(salary),
    sd_salary = sd(salary)
  )
```

\newpage

## What's the point?

Mean, Variance, and Standard deviation are sensitive to outliers and skewness.  They should only be used when a distribution looks "nice" (unimodal, symmetric, no outliers).  Otherwise, use median and IQR to summarize center and spread.


\begin{tabular}{p{5cm} p{6cm} p{6cm}}
\toprule
If the Distribution is... & Summarize Center with... & Summarize Spread with... \\
\midrule
Unimodal, Symmetric, and no Outliers & mean most common; median also OK & standard deviation most common; variance or IQR also OK \\
\midrule
Multimodal or Skewed or has Outliers & median & IQR \\
\bottomrule
\end{tabular}
