---
title: "R Commands for Quantitative Variables"
author: ''
date: "September 17, 2018"
output:
  pdf_document:
    keep_tex: true
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
   - \usepackage{booktabs}
documentclass: extarticle
classoption: 14pt
geometry: margin=0.6in
---


```{r, include=FALSE, message = FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
library(readr)
library(dplyr)
require(mosaic)
library(grid)
```

```{r, include=FALSE, message = FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)

require(fivethirtyeight)                
```


## Birth Weights and Tobacco Use During Pregnancy

Are babies' birth weights affected by whether or not the mother used tobacco during pregnancy?  Low birth weights are a risk factor for health problems later in life, so this is important!

Here is a data set with data on a sample of randomly selected babies who were born in North Carolina in 2004, with some information about the mother and the babies' weights in grams:

```{r, echo = FALSE, message = FALSE}
library(resampledata)
NCBirths2004 <- NCBirths2004[, 1:7]
```

```{r}
dim(NCBirths2004)
head(NCBirths2004)
```

Here's a plot of the data:

```{r, echo = FALSE}
ggplot(data = NCBirths2004, mapping = aes(x = Weight, color = Tobacco)) +
  geom_density()
```

### Warm Up #1: What did the code to make that plot look like?  Fill in the blanks below.

There are 4 blanks: what was used for data? For the aesthetic mapping to x?  For the aesthetic mapping to color?  For the geometry?

```{r, echo = TRUE, eval = FALSE}
ggplot(data =                  ,
    mapping = aes(x =                , color =                 )) +
  geom_                   ()
```

\vspace{1cm}

### Warm Up #2: What statistics would you use to summarize the center and spread of the distribution of birth weights within each group?

\newpage

## Calculating Summaries of Quantitative Variables: `summarize` and `group_by`

We saw from the plot that there seems to be a difference in birth weights.  What are "typical" birth weights for the two groups?

```{r}
NCBirths2004 %>%
  group_by(Tobacco) %>%
  summarize(
    mean_wt = mean(Weight),
    median_wt = median(Weight),
    q1_wt = quantile(Weight, probs = 0.25),
    q3_wt = quantile(Weight, probs = 0.75),
    iqr_wt = IQR(Weight),
    var_wt = var(Weight),
    sd_wt = sd(Weight)
  )
```

### Note: if we wanted to just calculate these summaries for all babies combined (across both groups), we would eliminate the `group_by` command:

```{r}
NCBirths2004 %>%
  summarize(
    mean_wt = mean(Weight),
    median_wt = median(Weight),
    q1_wt = quantile(Weight, probs = 0.25),
    q3_wt = quantile(Weight, probs = 0.75),
    iqr_wt = IQR(Weight),
    var_wt = var(Weight),
    sd_wt = sd(Weight)
  )
```

\newpage

## Sorting the data: `arrange`

### What was the shortest gestation time?

To answer this question, we will `arrange` the babies in increasing order of gestation time.  Then, the babies with the shortest gestation times will be near the `head` of the data frame.

```{r}
NCBirths_by_gestation <- NCBirths2004 %>%
  arrange(Gestation)

head(NCBirths_by_gestation)
```

\vspace{1cm}

### What was the longest gestation time?

Now we want to `arrange` the babies in `desc`ending order of gestation time, so the longest gestation times will be at the `head` of the data frame:

```{r}
NCBirths_by_gestation_descending <- NCBirths2004 %>%
  arrange(desc(Gestation))

head(NCBirths_by_gestation_descending)
```

\vspace{1cm}

\newpage

## Modifying or Adding a Variable: `mutate`

The babies' birth `Weights` are currently recorded in grams.  Let's add a new variable to the data frame called `Weight_lbs` with the babies' birth weights in pounds.

There are about 453.6 grams in a pound.  We can divide the weight in grams by 453.6 to get the weight in pounds.

```{r}
NCBirths2004 <- NCBirths2004 %>%
  mutate(
    Weight_lbs = Weight / 453.6
  )
```

As usual, we can take a look at the results by looking at the output from `head` or `str`.

```{r}
head(NCBirths2004)
```


